\mythesistitle

### 2011-12-21 src/thesis.tex
Updated thesis and cop presentaiton files to sync up with mylatexlib.sty. modified latter for the same as well.
Fixed the font issue in presentation which had also manifested in synopsis+thesis. Package txfonts was the culprit.
cleaned up library - new sty for thesis files. now Beamer and thesis files can coexist on mylatexlib.sty.

\documentclass[synopsis,MS]             % MS thesis dissertation.
              {iitmdiss}

\usepackage{times}             % Font?
\usepackage{t1enc}             % TI encoding??  
                            %%% Graphics related.                 %

\usepackage{../lib/mylatexlib}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title page
\title{\mythesistitle}

\author{\myname}

\date{\MakeUppercase {\mysubmissionmonth}}
\department{\MakeUppercase {\mydept}}

@@

\begin{singlespacing}
  \hspace*{-0.25in}
  \parbox{2.5in}{
    \noindent {\bf Prof.~\myadvisor} \\
    \noindent Research Guide \\
    \noindent Associate Professor \\
    \noindent Dept. of \mydept\\
    \noindent IIT Madras -- 600 036} \hfill
  \parbox{1.25in}{               % {1.75in}
    \vspace*{0.75in} 
    \noindent Chennai\\          %{\em Place:}
    \noindent \mysubmissiondate} %{\em Date:}
\end{singlespacing}

@@

\tnote[a]{work in progress.}

@@

\def \tem {}
  \noindent 
  \annote[a]{In this paper, we explore a natural generalization of
  results on binary matrices with the |\tem consecutive ones property|. 
  We consider the following constraint satisfaction problem. Given (i)
  a set system $\F \subseteq$ $(2^{U} \setminus \emptyset)$ of a
  finite set $U$ of cardinality $n$, (ii) a tree $T$ of size $n$ and
  (iii) a bijection called |\tem tree path labeling|, $\cl$ mapping the
  sets in $\cF$ to paths in $T$, does there exist at least one
  bijection $\phi:U \rightarrow V(T)$ such that for each $S \in \cF$,
  $\{\phi(x) \mid x \in S\} = \cl(S)$?  A tree path labeling of a set
  system is called |\tem feasible| if there exists such a bijection
  $\phi$.  We present an algorithmic characterization of feasible tree
  path labeling. COP is a special instance of tree path labeling
  problem when $T$ is a path.  We conclude with a polynomial time
  algorithm to find a feasible tree path labeling of a given set
  system when $T$ is a |\tem $k$-subdivided star|, set system has a
  single containment tree of overlap components and set size is
  limited to at most $k+2$. }{this is the abstract of the
  paper. expand to fit thesis.}
\end{abstract}

@@

\noindent 
\begin{tabbing}
xxxxxxxxxxx \= xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx \kill
\textbf{COP}   \> Consecutive Ones Property \\
\textbf{COT}   \> Consecutive Ones Testing \\
\textbf{ICPIA}   \> Intersection Cardinality Preservation Interval Assignment \\
\textbf{ICPPL}   \> Intersection Cardinality Preserved Path Labeling \\

\end{tabbing}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Introduction.

\chapter{Introduction}
\label{chap:intro}
\tnote[a]{ Some things to remember - have this somewhere.

>>>>

\documentclass[MS]             % MS thesis dissertation.
              {iitmdiss}

\usepackage{times}             % Font?
\usepackage{t1enc}             % T1 encoding??  
%\usepackage{boxedminipage}
\usepackage{../lib/mythesislib}
\usepackage{../lib/mylatexlib}

\def \mythesissubmissiondate {30 December 2011}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title page
\title{\mythesistitle} 
\author{\myname} 
\date{\MakeUppercase
  {\mythesissubmissionmonth}} 
\department{\MakeUppercase{\mydept}}

@@

\begin{singlespacing}
  \hspace*{-0.25in}
%   \begin{tabular}[h]{lr}
%     {\bf Dr.~\myadvisor} & \\
%     {\small Research Guide} & \\
%     {\small Associate Professor} & \\
%     {\small Dept. of \mydept} & \hfill {\small Chennai}\\
%     {\small IIT Madras -- 600 036} & \hfill {\small\mythesissubmissiondate}
%  \end{tabular}
  \parbox{3in}{
    \noindent {\bf Dr.~\myadvisor} \\
    \small
    \noindent Research Guide\\
    \noindent Associate Professor\\
    \noindent Dept. of \mydept\\
    \noindent IIT Madras -- 600 036
  } \hfill
  \parbox{1.25in}{               % {1.75in}
    \small
    \vspace*{0.75in}
    \begin{tabular}[h]{r}
    \noindent {Chennai}\\         
    \noindent {\mythesissubmissiondate} 
    \end{tabular}
  }
\end{singlespacing}

@@

\tnote{WIP}

@@

\def \tem {}
  \noindent 
  \annote{In this paper, we explore a natural generalization of
  results on binary matrices with the |\tem consecutive ones property|. 
  We consider the following constraint satisfaction problem. Given (i)
  a set system $\F \subseteq$ $(2^{U} \setminus \emptyset)$ of a
  finite set $U$ of cardinality $n$, (ii) a tree $T$ of size $n$ and
  (iii) a bijection called |\tem tree path labeling|, $\cl$ mapping the
  sets in $\cF$ to paths in $T$, does there exist at least one
  bijection $\phi:U \rightarrow V(T)$ such that for each $S \in \cF$,
  $\{\phi(x) \mid x \in S\} = \cl(S)$?  A tree path labeling of a set
  system is called |\tem feasible| if there exists such a bijection
  $\phi$.  We present an algorithmic characterization of feasible tree
  path labeling. COP is a special instance of tree path labeling
  problem when $T$ is a path.  We conclude with a polynomial time
  algorithm to find a feasible tree path labeling of a given set
  system when $T$ is a |\tem $k$-subdivided star|, set system has a
  single containment tree of overlap components and set size is
  limited to at most $k+2$. }{this is the abstract of the
  paper. expand to fit thesis.}
\end{abstract}

@@

\noindent 
\begin{tabbing}
xxxxxxxxxxx \= xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx \kill
\textbf{COP}   \> Consecutive Ones Property \\
\textbf{COT}   \> Consecutive Ones Testing \\
\textbf{ICPIA}   \> Intersection Cardinality Preservation Interval Assignment \\
\textbf{ICPPL}   \> Intersection Cardinality Preserved Path Labeling \\
% ADD MORE HERE
\end{tabbing}

@@

\chapter{Introduction}
\label{chap:intro}

\section{Introduction}
\label{sec:intro}
\vspace{\secfirstparatrim}

Consecutive ones property is a non-trivial property of binary matrices
that has been studied widely in the literature for over past 50
years. Detection of COP in a matrix is possible efficiently and there
are several algorithms that achieve the same. This thesis documents
the work done on an extension of COP extended from the equivalent
interval assignment problem in \cite{nsnrs09}. These new results
rigorously prove a natural extension (to trees) of their
characterization as well as makes connections to graph isomorphism,
namely path graph isomorphism.

% \tnote{testing 1 2 3...}
% \tnote[bogus1]{testing 1 2 3...}
% \tnote[bogus2]{testing 1 2 3...}
% \tnote[GTC]{testing 1 2 3...}
% \tnote[bogus4]{testing 1 2 3...}
% \tnote[TH]{testing 1 2 3...}

\subsection{Illustration of the problem}
\label{sec:problem}
\vspace{\secfirstparatrim}

A group of students, \Pa, \Pi, \Sn, \Wo, \Vi, \Li, \Ch, \Sa, \Fr,
  \Sc\ and \Lu\ enroll at the
% {\bf Wo}odstock, {\bf Vi}olet, {\bf Li}nus, {\bf Ch}arlie, {\bf
%   Sa}lly, {\bf Re}run, {\bf Sc}hr{\"o}eder} and {\em {\bf Lu}cy}
% enroll at the
  {\WSI} for a liberal arts programme.  As part of their semester
  thesis, they pick a body of work to study and form the namesake
  study groups, {\LLL}, {\GGG}, {\BBB} and {\TTT}\tnote[TH]{put bib
    entries for these works!}. A student will be in at least one study
  group and may be in more than one. For instance, as will be seen
  later, {\Fr} studies both {\LLL} and {\TTT} while \Wo\ studies only
  \BBB.

Let $U$ and $\cF$ represent the set of students and the set of study
groups respectively and the integers $n$ and $m$ denote the total
number students and study groups respectively. In relation to this
example, these are defined in Table~\ref{tab:wsigroups}. Also given
there is the study group allocation to students.

% \vspace{-3mm}
% \begin{align*}
% U &= \{\xPa,\; \xPi,\; \xSn,\; \xWo,\; \xVi,\; \xLi,\; \xCh,\;
%     \xSa,\; \xFr,\; \xSc,\; \xLu\}\\
% \cF &= \{\xLLL, \xGGG, \xBBB, \xTTT\}\\
% \xLLL &= \{\xCh,\;  \xSa,\;  \xFr,\;  \xSc,\;  \xLu \}\\
% \xGGG &= \{\xPa,\;  \xPi,\;  \xVi,\;  \xCh \}\\
% \xBBB &= \{\xSn,\;  \xPi,\;  \xWo \}\\
% \xTTT &= \{\xVi,\;  \xLi,\;  \xCh,\;  \xFr \}\\
% n &= |U| = 11\\
% m &= |\cF| = 4  
% \end{align*}

\begin{table}[htbp]
  \centering
  { \footnotesize
  \begin{tabular}{rcl}
    $U $&$=$&$ \{\xPa,\; \xPi,\; \xSn,\; \xWo,\; \xVi,\; \xLi,\; \xCh,\;
    \xSa,\; \xFr,\; \xSc,\; \xLu\}$\\
    $\cF $&$=$&$ \{\xLLL, \xGGG, \xBBB, \xTTT\}$\\
    $\xLLL $&$=$&$ \{\xCh,\;  \xSa,\;  \xFr,\;  \xSc,\;  \xLu \}$\\
    $\xGGG $&$=$&$ \{\xPa,\;  \xPi,\;  \xVi,\;  \xCh \}$\\
    $\xBBB $&$=$&$ \{\xSn,\;  \xPi,\;  \xWo \}$\\
    $\xTTT $&$=$&$ \{\xVi,\;  \xLi,\;  \xCh,\;  \xFr \}$\\
    $n $&$=$&$ |U| = 11$\\
    $m $&$=$&$ |\cF| = 4$      
  \end{tabular}
}
  \caption{\figtabsize Students and study groups in \WSI}
  \label{tab:wsigroups}
\end{table}

% The administration faces a small challenge with their
% accommodation. 
The campus has a residential area {\residenceblock} that has $n$
single occupancy apartments reserved for the study groups'
accommodation.  All these apartments are located such that the streets
connecting them do {\em not} form
loops. Fig~\ref{fig:streetmappathpeople} \tnote[TH]{fig:streetmap}
shows the street map for {\residenceblock}. It may be noted that as a
graph, it classifies as a tree.

\begin{table}[htbp]
  \centering
  {\footnotesize    
    \begin{tabular}{l||l}
      \begin{tabular}{rcl}
        $T $&$=$&$ \text{\em Street map tree of \residenceblock}$\\
        $V(T) $&$=$&$ \{ 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11 \}$\\
        % $\phi^{-1}(V(T)) $&$=$&$ \{ \xSa,\; \xPi,\; \xFr,\; \xWo,\;
        % \xCh,\; \xVi,\;
        % \xPa,\; \xSn,\; \xLu,\; \xLi,\; \xSc \}$\\
        $\cP $&$=$&$ \{R\xLLL, R\xGGG, R\xBBB, R\xTTT\}$\\
        $R\xLLL $&$=$&$ \{9, 1, 5, 3, 11\}$\\
        $R\xGGG $&$=$&$ \{7, 2, 6, 5\}$\\
        $R\xBBB $&$=$&$ \{8, 2, 4 \}$\\
        $R\xTTT $&$=$&$ \{10, 6, 5, 3\}$\\
        $n $&$=$&$ |V| = 11$\\
        $m $&$=$&$ |\cP| = 4$\\\\
        $\cl $&$=$& {\em Study group to route mapping}\\        
        $\cl(\mathbb{X}) $&$=$&  $R\mathbb{X}$ for all $\mathbb{X} \in
        \cF$
      \end{tabular} & 
      \begin{tabular}{c}
       {\em Apartment allocation }($\phi$)\\
                    \begin{tabular}{c|c}                      
                      1  & \xSa\\
                      2  & \xPi\\
                      3  & \xFr\\
                      4  & \xWo\\
                      5  & \xCh\\
                      6  & \xVi\\
                      7  & \xPa\\
                      8  & \xSn\\
                      9  & \xLu\\
                      10 & \xLi\\
                      11 & \xSc                     
                    \end{tabular}
      \end{tabular}\\
    \end{tabular}
  }
  \caption{\figtabsize A solution to study group accommodation problem}
  \label{tab:iltree}
\end{table}

A natural question would be to find how the students should be
allocated apartments such that each study group has the {\em least
  distance to travel} for a discussion? More specifically, we are
interested in enforcing additional conditions, namely, that all the
students in a study group must be next to each other; in other words,
for one student to reach another fellow study group member's apartment
(for all study groups the student is part of), she must not have to
pass the apartment of any student who is not in that study group. To
further elucidate, the apartments of students of any study group must
be arranged in an exclusive unfragmented path on the street
map. Exclusivity here means that the path must not have apartments
from other study groups (unless that apartment is also part of {\em
  this} study group).

An intuitive approach to this problem would be to first find the paths
that each study group decides to inhabit and then refine the
allocation to individual students. A feasible allocation of exclusive
routes to study groups is illustrated in
Fig~\ref{fig:streetmappathpeople} \tnote[TH]{fig:streetmappath} and
the students' allocation of apartments that obeys this route
allocation is also shown. Table~\ref{tab:iltree} shows the same
solution set theoretically.  How this is algorithmically computed is
the focus of this thesis.

%%%%%%%%%%%% THESIS UNCOMMENT %%%%%%%%%%%%
% \begin{figure}[t]%[htbp]
%  \centering
%   \includegraphics[scale=0.3]{../img/1_infinite_loop.png}
%   \caption{{\residenceblock} street map}
%   \label{fig:streetmap}
% \end{figure}

% \begin{figure}[t]%[htbp]
%   \centering
%   \includegraphics[scale=0.3]{../img/2_infinite_loop_BTWF.png}
%   \caption{{\residenceblock} street map with study group routes
%     allocated. Routes are color coded as follows: red for
%     \textcolor{red}{$\xLLL$} group, blue for \textcolor{blue}{$\xGGG$}
%     group, orange for \textcolor{YellowOrange}{$\xBBB$} group, green for
%     \textcolor{green}{$\xTTT$} group}
%   \label{fig:streetmappath}
% \end{figure}
% \begin{figure}[t] % [h]
%   \centering
%   \includegraphics[scale=0.5]{../img/3_infinite_loop.png}
%   \caption{Individual allocation of apartments to students in
%     {\residenceblock} that meets the requirements stated before.}
%   \label{fig:streetmappathpeople}
% \end{figure}
%%%%%%%%%%%% END THESIS UNCOMMENT %%%%%%%%%%%%
% \begin{figure}[htbp] %[t]%
%  \centering
%   (a) \includegraphics[scale=0.2]{../img/1_infinite_loop.png}
%   (b) \includegraphics[scale=0.2]{../img/2_infinite_loop_BTWF.png}
%  \label{fig:streetmap}
%   \caption{\figtabsize (a) {\residenceblock} street map (b) {\residenceblock} street map with study group routes
%     allocated. Routes are color coded as follows: red for
%     \textcolor{red}{$\xLLL$} group, blue for \textcolor{blue}{$\xGGG$}
%     group, orange for \textcolor{YellowOrange}{$\xBBB$} group, green for
%     \textcolor{green}{$\xTTT$} group}
% \end{figure}
\begin{figure}[htbp] %[t]%
  \centering
  \includegraphics[scale=0.3]{../img/3_infinite_loop.png}
  \caption{\figtabsize Individual allocation of apartments to students
    in {\residenceblock} that meets the requirements stated before.
    The routes are color coded as follows: red for
    \textcolor{red}{$\xLLL$} group, blue for \textcolor{blue}{$\xGGG$}
    group, orange for \textcolor{yellow}{$\xBBB$} group, green
    for \textcolor{green}{$\xTTT$} group. {\tiny {\em Peanuts images
        {\copyright} Charles Schulz}}}% YellowOrange
  \label{fig:streetmappathpeople}
\end{figure}
\tnote[TH]{make dashed/textured lines for routes. make it color agnostic.}

As a special case, suppose all the apartments are on the same street
or if they are all lined up on a single path, the street map becomes a
tree that is just a path. Then the problem becomes what is called an
{\em interval assignment problem}. The idea of interval assignment may
not be obvious here; hence to see this, consider a different problem
in {\WSI} where the classes for these study groups courses need to be
scheduled during a day (or a week or any time period). Each study
group has a bunch of courses associated with it some of which may be
shared by two or more study groups. It is mandatory that a student who
is a member of a study group takes all the courses associated with
that group. There are slots during the day for classes to be held and
the problem is to allocate class slots to courses such that all the
classes of a study group are consecutive. It is debatable if this will
not hamper the attention span and memory retention rate of the
students but that is, regrettably, out of the scope of this
thesis. The parallels between this class allocation problem and the
accommodation problem can be seen as follows. The set $U$ here, are
the courses offered (say Course 101 {\coneohone}, Course 102
{\coneohtwo} and so on). In this variation of the problem, the
collection $\cF$ is the set of study groups but the study groups are
filled by course IDs (in place of students in the earlier
example). For instance, Course 101 is mandatory for all study groups
$\xLLL$, $\xGGG$, $\xBBB$, $\xTTT$ and Course 102 is mandatory for
only the $\xLLL$ group) and so on. The sequence of class slots for the
day (or week or any time period) is analogous to the street map in the
accommodation problem. It is quite obvious now why this version of the
problem (where the ``target graph'' is a path and not any
tree\tnote[TH]{Allowing any tree in this example could be seen as a
  scenario where there are parallel classes. A node falling in the
  path between two other nodes would mean that the corresponding is
  scheduled between the other two.}) is called an interval assignment
problem.

The interval assignment problem to a set system is equivalent to the
consecutive ones property (COP) problem in binary matrices\cite{wlh02,
  nsnrs09}.  The COP problem is to rearrange rows (columns) of a
binary matrix in such a way that every column (row) has its {\un}s
occur consecutively. If this is possible the matrix is said to have
the COP.  COP is a well researched combinatorial problem and has
several positive results on tests for it and computing the COP
permutation (i.e. the course schedule in the above illustration) which
will be surveyed later in this document. Hence we are interested in
extensions of COP, more specifically, the extension of interval
assignment problem to tree path assignment problem (which is
illustrated by the study group accommodation problem).

Section~\ref{sec:background} gives a brief survey of COP and
optimization problems related to it followed by motivation for the
thesis in Section~\ref{sec:motive}.  Section~\ref{sec:results}
presents a summary of our results on the extension of COP namely, the
tree path labeling problem. \tnote[TH]{update for thesis}

\section{Consecutive Ones Property Testing - a Survey} %{Background}
\label{sec:background}
\vspace{\secfirstparatrim}
In this section, a brief survey of the consecutive ones problem and its
optimization problems is presented.

%%%%%%                              %%%%%%
%THX is the second level of ``tnote''-ing%
%%%%%%                              %%%%%%
\tnote[TH]{ADD: As it will be described in detail later in this
  document, isomorphism of certain % classes of graphs, namely chordal
%   graphs, have a close relationship with consecutive ones property and
%   generalizations of it.  This is perhaps because of how closely COP
%   of a matrix relates to properties of graphs derived from matrices as
%   seen in the following results.
}
\tnote[THX]{ADD: peo exists iff chordal. lexicographic BFS
  [tag:chordalGraph]} %
\tnote[THX]{ADD: A well known result in %perfect graph theory is that
%  the maximal cliques of  an interval graph $G$ can be linearly ordered
%   such that for all $v \in V(G)$, cliques containing $v$ are
%   consecutive in the ordering cite~{gh64}. This clearly means that a
%   graph $G$ is an interval graph if and only if
}% 
\tnote[THX]{verify
  from paper the statement of claim.}% 
\tnote[TH]{maximal clique vertex
  incidence matrix of%  $G$ has COP.  Also, maximal cliques of any
%   chordal graph can be enumerated in polytime $O(m+n)$
}
\tnote[THX]{citation?!!}%  
\tnote[TH]{cite~{fg65} uses these results to
  give the first polynomial time algorithm for COT.}%
\tnote[THX]{check. how do they use it?}%  
\tnote[TH]{ A bipartite graph
  is convex%  (on $R$) if and only if its half adjacency matrix has COP
%   on rows.  The results in cite~{bl76} on COT are based on the result
%   that interval graphs are AT-free chordal graphs.
}%
\tnote[THX]{the latter being Tucker's?}%  
\tnote[THX]{(2) TBD survey
  -- % see blue notes (in notebook) under Graph Isomorphism. namely
%   citations in cite:aas93 (3) a brief on heirarchy of chordal graphs,
%   path graphs, interval graphs, peo, clique tree etc. and results. (4)
%   interval graphs are incomparability graphs. see
%   Golumbic. [tag:classification] (5) finding min length hole in
%   bipartite graph is polynomial time [Sec 3.3.2 in cite:d08phd] (6)
%   theorem 2.2 in cite:d08phd - $G$ is union of v.d. caterpillars iff
%   $M$ has COP in rows. $M$ is edge vertex incidence matrix $(*,2)$ of
%   $G$
}%

\subsection{Matrices with COP}
\label{sec:copmatrices}
\vspace{\secfirstparatrim} 
As seen earlier, the interval assignment problem (illustrated as the
course scheduling problem in Section~\ref{sec:problem}), is a special
case of the problem we address in this thesis, namely the tree path
labeling problem (illustrated as the study group accommodation
problem). The interval assignment problem and COP problem are equivalent
problems. In this section we will see some of the results that
exists in the literature today towards solving the COP problem and
optimization problems surrounding it.

\begin{figure}[t] %[htbp]
  \centering
  % \includegraphics[scale=0.4]{../img/0_cop_nocop.png}
{\figtabsize
 \begin{tabular}[h]{l|lcccl}
    $M_1$: & $M_1'$: &&&& $M_2$:\\
    &&&&&\\
    \begin{tabular}[h]{llll}
      $c_1$ & $c_2$ &$c_3$ &$c_4$\\
      &&&\\
      \un & 0   & \un & 0\\
      0   & \un & 0   & \un \\
      \un & 0   & 0   & \un
    \end{tabular}
    &
    \begin{tabular}[h]{llll}
      $c_3$ &$c_1$ &$c_4$& $c_2$\\
      &&&\\
      \un & \un & 0 & 0\\
      0 & 0 & \un & \un \\
      0 & \un & \un & 0
    \end{tabular}
    &&&&
    \begin{tabular}[h]{llll}
      $d_1$ & $d_2$ &$d_3$ &$d_4$\\
      % &&&\\
      &&&\\
      \un & \un & 0 & 0\\
      0 & \un & \un & 0 \\
      0 & \un & 0 & \un 
    \end{tabular}
  \end{tabular}
}

  \caption{\figtabsize Matrices with and without COP. $M_1$ has COP because by
    permuting its columns, $c_1$-$c_4$, one can obtain $M_1'$ where
    the {\un}s in each row are consecutive. $M_2$, however, does not
    have COP since no permutation of its columns, $d_1$-$d_4$, will
    arrange {\un}s in each row consecutively \cite{d08phd}.}

  \label{fig:cop-matrix}
\end{figure}

Recall that a matrix with COP is one whose rows (columns)
can be rearranged so that the {\un}s in every column (row) are in
consecutive rows (columns). Figure~\ref{fig:cop-matrix} shows examples
of this property.
COP in binary matrices has several practical applications in diverse
fields including scheduling \cite{hl06}, information retrieval
\cite{k77} and computational biology \cite{abh98}.  Further, it is a
tool in graph theory \cite{mcg04} for interval graph recognition,
characterization of Hamiltonian graphs, planarity testing \cite{bl76}
and in integer linear programming \cite{ht02,hl06}.

The obvious first questions after being introduced to the consecutive
ones property of binary matrices are if COP can be detected
efficiently in a binary matrix and if so, can the COP permutation of
the matrix also be computed efficiently?  Recognition of COP in a
binary matrix is polynomial time solvable and the first such algorithm
was given by \cite{fg65}.  A landmark result came a few years later
when \cite{at72} discovered the families of forbidden submatrices that
prevent a matrix from having COP and most, if not all, results that
came later were based on this discovery which connected COP in binary
matrices to convex bipartite graphs. In fact, the forbidden
submatrices came as a corollary to the discovery that convex bipartite
graphs are AT-free in \cite{at72}\tnote[GTC]{check}. The first linear time
algorithm for COP testing (COT) was invented by \cite{bl76} using a
data structure called PQ trees.  Since then several COT algorithms
have been invented -- some of which involved variations of PQ trees
\cite{mm96,wlh01,mcc04}, some involved set theory and ICPIA
\cite{wlh02,nsnrs09}, parallel COT algorithms\cite{as95,bs03,ly91} and
certifying algorithms\cite{mcc04}. % These algorithms
% improve over PQ trees on the time complexity and/or the complexity
% of implementation.

The construction of PQ trees in \cite{bl76} draws on the
close relationship of matrices with COP to interval graphs. A PQ tree of a
matrix is one that stores all row (column) permutations of the matrix
that give the COP orders (there could be multiple orders of rows or columns) of the matrix. This
is constructed using an elaborate linear time procedure and is also
a test for planarity\tnote[GTC]{check check check. both interval
  graph and planarity in this paper?}. % Since this
% result, several improvements have been made to the procedural
% complexity of the algorithm itself and also derivative data structures
% that are easier to construct.
PQR trees is a generalized data structure based on PQ trees
\cite{mm96,mpt98}.  \cite{tm05} describes an improved algorithm to
build PQR trees. \tnote[GTC]{improv in terms of what?}\cite{wlh02}
describes the simpler algorithm for COT. Hsu also invented PC trees
\cite{wlh01}\tnote[TH]{This result first appeared inproc ISAAC92}
which is claimed to be much easier to implement. \cite{nsnrs09}
describes a characterization of consecutive ones property solely based
on the cardinality properties of the set representations of the
columns (rows); every column (row) is equivalent to a set that has the
row (column) indices of the rows (columns) that have one entries in
this column (row). This is interesting and relevant, especially to
this thesis because it simplifies COT to a great
degree. \tnote[GTC]{it reduces the solution search space. fill in the blanks.}

\cite{mcc04} describes a different approach to COT. While all previous
COT algorithms gave the COP order if the matrix has the property but
exited stating negative if otherwise, this algorithm gives an evidence
by way of a certificate of matrix even when it has no COP. This
enables a user to verify the algorithm's result even when the answer
is negative. This is significant from an implementation perspective
because automated program verification is hard and manual verification
is more viable. Hence having a certificate reinforces an
implementation's credibility. Note that when the matrix {\em has} COP,
the COP order is the certificate.  The internal machinery of this
algorithm is related to the weighted betweenness problem
addressed\tnote[GTC]{in what way??} in \cite{co98}.  \tnote[GTC]{expand
  on the COP order graph creation and it having to be bipartite for M
  to have COP. and thus an odd cycle being an evidence of no COP.}
\tnote[TH]{ where should this go?: (1) cite|jlm97 (application of PQ trees
  in graphics). (2) helly's theorem citation
  19XXdgk-Hellystheorem-Danzer-Gruenbaum-Klee}

\subsection{Optimization problems in COP}
\label{sec:optcop}
\vspace{\secfirstparatrim}  
So far we have been concerned about matrices that have the consecutive
ones property. However in real life applications, it is rare that data
sets represented by binary matrices have COP, primarily due to the
noisy nature of data available. At the same time, COP is not arbitrary
and is a desirable property in practical data representation
\cite{co98,jkckv04,k77}. In this context, there are several
interesting problems when a matrix does not have COP but is ``close''
to having COP or is allowed to be altered to have COP. These are the
optimization problems related to a matrix which does not have
COP. Some of the significant problems are surveyed in this section.

\tnote[TH]{ -- sect 4.1 in cite:d08phd has many results
  surveyed. hardness results, approx. results. results are usually for
  a class of matrices $(a,b)$ where number {\un}s in columns and rows
  are restriced to $a$ and $b$ . -- problem of flipping at most $k$
  entries of $M$ to make it attain COP. this is NP complete
  cite:b75-phd}\tnote[TH]{(1) scite:lb62 showed that interval graphs
  are AT-free.  describe AT (2) show the close relationship b/w COP
  and graphs sec 2.2, pg 31} \cite{at72} showed that a matrix that
does not have COP have certain substructures that prevent it from
having COP. Tucker classified these forbidden substructures into five
classes of submatrices. This result is presented in the context of
convex bipartite graphs which \cite{at72} proved to be
AT-free\tnote[TH]{ check this up. give details. - doms'}. By
definition, convex bipartite graph have half adjacency matrices that
have COP on either rows or columns (graph is biconvex if it has COP on
both)\cite{d08phd}. A half adjacency matrix is a binary matrix
representing a bipartite graph as follows. The set of rows and the set
of columns form the two partitions of the graph. Each row node is
adjacent to those nodes that represent the columns that have {\un}s in
the corresponding row. \cite{at72} proves that this bipartite graph
has no asteroidal triple if and only if the matrix has COP and goes on
to identify the forbidden substructures for these bipartite
graphs. The matrices corresponding to these substructures are the
forbidden submatrices.

Once a matrix has been detected to not have COP (using any of the COT
algorithms mentioned earlier), it is naturally of interest to find out
the smallest forbidden substructure (in terms of number of rows and/or
columns and/or number of entries that are {\un}s). \cite{d08phd}
discusses a couple of algorithms which are efficient if the number of
{\un}s in a row is small. This is of significance in the case of sparse
matrices where this number is much lesser than the number of
columns. $(*,\Delta)${\em -matrices} are matrices with no restriction
on number of {\un}s in any column but has at most $\Delta$ {\un}s in
any row. {\sc Min COS-R (Min COS-C), Max COS-R (Max COS-C)} are
similar problems which deals with inducing COP on a matrix. In {\sc
  Min COS-R (Min COS-C)} the question is to find the minimum number of
rows (columns) that must be deleted to result in a matrix with COP.
In the dual problem {\sc Max COS-R (Max COS-C)} the search is for the
maximum number of rows (columns) that induces a submatrix with
COP. Given a matrix $M$ with no COP, \cite{b75-phd} shows that finding
a submatrix $M'$ with all columns\tnote[GTC]{check if b75 deals with
  COP col or COP row. also is it any submatrix with k less than r rows
  or submatrix must have all columns?} but a maximum cardinality
subset of rows such that $M'$ has COP is NP complete. \cite{hg02}
corrects an error of the abridged proof of this reduction as given in
\cite{gj79}.  \cite{d08phd} discusses all these problems in detail
giving an extensive survey of the previously existing results which
are almost exhaustively all approximation results and hardness
results. Taking this further, \cite{d08phd} presents new results in
the area of parameterized algorithms for this
problem\tnote[TH]{elaborate - what are the results?}.

Another problem is to find the minimum number of entries in the matrix
that can be toggled to result in a matrix with COP.  \cite{v85}
discusses approximation of {\sc COP Augmentation} which is the problem
of changing of the minimum number of zero entries to {\un}s so that the
resulting matrix has COP. As mentioned earlier, this problem is
known to be NP complete due to \cite{b75-phd}. \cite{v85} also proves,
using a reduction to the longest path problem, \tnote[GTC]{or is it a
  survey of another result?  check.} that finding a Tucker's forbidden
submatrix of at least $k$ rows is NP complete. \tnote[GTC]{how is this
  different from booth's 75 result??}  \tnote[TH]{where should this
  go? cite|tz04 (approx submatrix with COP sparse matrices)}

\cite{jkckv04} discusses the use of matrices with almost-COP (instead
of one block of consecutive {\un}s, they have $x$ blocks, or {\em
  runs}, of consecutive {\un}s and $x$ is not too large) in the
storage of very large databases.  The problem is that of reordering of
a binary matrix such that the resulting matrix has at most $k$ runs of
{\un}s. This is proved to be NP hard using a reduction from the
Hamiltonian path problem.\tnote[TH]{Theorem 2.1 in jkckv}
\tnote[TH]{(1) A connection of COP problem to the travelling salesman
  problem is also introduced. what does this mean? -- COP can be used
  as a tool to reorder $0.5T \le runs(M) \le T$. (2) The optimization
version of the $k$-run problem, i.e. minimization of number of blocks
of ones is proven to be NP complete by cite:k77}\tnote[GTC]{are these two
  the same?} \tnote[TH]{what is the reduction?} \tnote[TH]{other
  problems similar to COP -- cite:ckl96 (ILP, circ ones, one drop) --
  cite:th98 (generalization of COP - minimax, biotonic column) Tucker}

\section{Generalization of COP - The Motivation}  %{Motivation}
\label{sec:motive}
\vspace{\secfirstparatrim}  
As seen in Section~\ref{sec:copmatrices}, \cite{nsnrs09} describes a
characterization of consecutive ones property based on the cardinality
properties of the set representations of the columns. This result is
very relevant to this thesis because aside from it simplifying COT to
a great degree, our generalization problem is motivated by their
results.

The result in
\cite{nsnrs09} characterizes interval assignments to the sets which
can be obtained from a single permutation of the rows.  They show that
for each set, the cardinality of the interval assigned to it must be
same as the cardinality of the set, and the intersection cardinality
of any two sets must be same as the intersection cardinality of the
corresponding intervals.  While this is obviously a necessary
condition, this result shows this is also sufficient.  \cite{nsnrs09}
calls such an interval assignment an Intersection Cardinality
Preserving Interval Assignment (ICPIA).  This paper generalizes the
idea from \cite{wlh02} of decomposing a given binary matrix into prime
matrices for COT and describes an algorithm to test if an ICPIA exists
for a given set system.

The tree path labeling problem is a natural generalization of the
interval assignment problem or the COP problem. The problem is defined
as follows -- given a set system $\cF$ from a universe $U$ and a tree
$T$, does there exist a bijection from $U$ to the vertices of $T$ such
that each set in the system maps to a path in $T$.  We refer to this
as the {\em tree path labeling problem} for an input set system,
target tree pair -- $(\cF,T)$. As a special case if the tree $T$ is a
path, the problem becomes the interval assignment problem.  We focus
on the question of generalizing the notion of an ICPIA \cite{nsnrs09}
to characterize feasible path assignments.  We show that for a given
set system $\cF$, a tree $T$, and an assignment of paths from $T$ to
the sets, there is a feasible bijection between $U$ and $V(T)$ if and
only if all intersection cardinalities among any three sets (not
necessarily distinct) is same as the intersection cardinality of the
paths assigned to them and the input passes a filtering algorithm
(described in this paper) successfully.  This characterization gives a
natural data structure that stores all the relevant feasible
bijections between $U$ and $V(T)$. This reduces the search space for
the solution considerably from the universe of all possible bijections
between $U$ and $V(T)$ to only those bijections that maintain the
characterization\tnote[TH]{REWORD}.  Further, the filtering algorithm
is also an efficient algorithm to test if a tree path labeling to the
set system is
feasible.  %This generalizes the result in \cite{nsnrs09}.
